*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-22 Code/Manuscript_O2.Rmd
editing introduction

## *Prochlorococcus* evolutionary and functional diversity {.unnumbered}

*Prochlorococcus marinus*, a genus of the phylum Cyanobacteria, is the smallest known photosynthetic prokaryote, with cell diameters ranging from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992]. Despite small cell size, *P. marinus* contribute 13 to 48% of net primary production in oligotrophic oceans, corresponding to about 30% of  global oxygen production [@partenskyProchlorococcusMarinePhotosynthetic1999].  *P. marinus* growth is currently constrained to between latitudes of 40°N to 40°S in open ocean waters, from surface to 300 m  depth, thus spanning 3 orders of magnitudes of light [@partenskyProchlorococcusMarinePhotosynthetic1999; @chisholmProchlorococcusMarinusNov1992 ].

*Prochlorococcus marinus* is functionally differentiated from other cyanobacteria primarily by the absence of phycobilisomes [@chisholmProchlorococcusMarinusNov1992]. Instead, *Prochlorococcus*  utilizes intra-membrane Prochlorophyte Chlorophyll Binding (Pcb) proteins bound to divinylchlorophyll a (Chl a2), and α-carotene, as the major light harvesting complex [@goerickePigmentsProchlorococcusMarinus1992]; [@larocheIndependentEvolutionProchlorophyte1996], with diversity in the number of genes encoding Pcb, expression of Pcb under changing light [@garczarekMultiplicationAntennaGenes2000]  [@rocapGenomeDivergenceTwo2003] [@garczarekMultiplicationAntennaGenes2000], and in thylakoid organization [@partenskyProchlorococcusMarinePhotosynthetic1999].  The genes encoding most of the rest of the photosynthetic apparatus is highly conserved in *Prochlorococcus* [@kettlerPatternsImplicationsGene2007], though the photosynthetic efficiency may vary among clades [@hessPhotosyntheticApparatusProchlorococcus2001; @moorePhotophysiologyMarineCyanobacterium1999].

*Prochlorococcus marinus* comprises many strains, organized into clades, defined by 16S-23S intergenic transcribed ribosomal sequence signatures (Rocap et al., 2003). The clades inhabit distinct ecological niches [@moorePhysiologyMolecularPhylogeny1998], originally defined by Low-Light (LL) or High-Light (HL).  Of at least 12 known *Prochlorococcus* clades only 5 to date have cultured representatives; HLI, HLII, LLI, LLII/III and LIV [@billerProchlorococcusStructureFunction2015].

XXXXMIREILLE Check strain descriptions for accuracyXXXX
Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral ranges [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, at depths from 120 m and 200 m (Partensky et al., 1999). LLIV members are, as yet, the only cultured strains found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @robledoCrypticOxygenCycling2017] (Johnson et al., 1999). *P. marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010].

Clades LLII and LLIII are often grouped together as the second oldest phylogenetic lineages in the *Prochlorococcus* radiation, and share an affinity for low light; LLIII cultured strain SS120, was isolated from 135 m depth in the Sargasso Sea.

Clade LLI  includes cultured strains NATL1A and NATL2A, which prefer moderate irradiances found between 30 and 100 m depth.

>>>>

*Prochlorococcus marinus*, a genus of Cyanobacteria, is the smallest known photosynthetic prokaryote, with cell diameters ranging from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992]. Despite small cell size, *P. marinus* contribute 13 to 48% of net primary production in oligotrophic oceans, corresponding to about 30% of global oxygen production [@partenskyProchlorococcusMarinePhotosynthetic1999].  *P. marinus* growth is currently constrained between latitudes of 40°N to 40°S in open ocean waters, from surface to 300 m depth, thus spanning 3 orders of magnitudes of light [@partenskyProchlorococcusMarinePhotosynthetic1999; @chisholmProchlorococcusMarinusNov1992 ].

*P. marinus* is functionally differentiated from other cyanobacteria primarily by the absence of phycobilisomes [@chisholmProchlorococcusMarinusNov1992]. Instead, *Prochlorococcus*  uses intra-membrane Prochlorophyte Chlorophyll Binding (Pcb) proteins binding divinylchlorophyll a (Chl a2), and α-carotene, as major light harvesting complexes [@goerickePigmentsProchlorococcusMarinus1992]; [@larocheIndependentEvolutionProchlorophyte1996], with diversity in the number of genes encoding Pcb, expression of Pcb under changing light [@garczarekMultiplicationAntennaGenes2000]  [@rocapGenomeDivergenceTwo2003] [@garczarekMultiplicationAntennaGenes2000], and in thylakoid organization [@partenskyProchlorococcusMarinePhotosynthetic1999].  The genes encoding much of the photosynthetic apparatus is highly conserved in *Prochlorococcus* [@kettlerPatternsImplicationsGene2007], though the photosynthetic efficiency may vary among clades [@hessPhotosyntheticApparatusProchlorococcus2001; @moorePhotophysiologyMarineCyanobacterium1999].

*P. marinus* comprises many strains, organized into clades, defined by 16S-23S intergenic transcribed ribosomal sequence signatures (Rocap et al., 2003). The clades inhabit distinct ecological niches [@moorePhysiologyMolecularPhylogeny1998], originally defined by Low-Light (LL) or High-Light (HL).  Of at least 12 known *Prochlorococcus* clades only 5 to date have cultured representatives; HLI, HLII, LLI, LLII/III and LIV [@billerProchlorococcusStructureFunction2015].

Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral range [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, typical of depths from 120 m to 200 m (Partensky et al., 1999). LLIV members are, as yet, the only cultured strains to have been found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI also thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @robledoCrypticOxygenCycling2017] (Johnson et al., 1999). *P. marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010], and may be net O~2~ consumers (2018, Partensky et al. 2018).

Clades LLII and LLIII are grouped together as the second oldest phylogenetic lineages diversifying in the *Prochlorococcus* radiation, and share an affinity for low light; LLIII cultured strain SS120, was isolated from 135 m depth in the Sargasso Sea.

Clade LLI  includes cultured strains NATL1A and NATL2A, which prefer moderate irradiances typical between 30 and 100 m depth.
